---
title: "Covid"
author: "B. Leblanc"
date: "2024-12-07"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

This is an R Markdown report on Covid 19 data across the US and globally.

## Initializing Data

We need to load all necessary libraries.

```{r library, message = FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
```

First I will read in the data from the csv files.

```{r data, message = FALSE}
## Get current data from files

url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c("time_series_covid19_confirmed_US.csv", "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv", "time_series_covid19_deaths_global.csv")
# concatenate the urls
urls <- str_c(url_in, file_names)
```

Now lets read in the data.

```{r import, message = FALSE}
global_cases <- read_csv(urls[2])
global_deaths <- read_csv(urls[4])
US_cases <- read_csv(urls[1])
US_deaths <- read_csv(urls[3])
```

## Tidy Data

After viewing the datasets, I would like to tidy global_cases and global_deaths and put each variable (date, cases, deaths) in their own column.
Also I don't need position variables (Lat, Long) for analysis so I will get rid of those and rename columns to be more R friendly.

```{r tidy_global_data, message = FALSE}

# pivot datset

global_cases <- global_cases %>%
  pivot_longer(cols = 
-c('Province/State', 'Country/Region', Lat, Long),
  names_to = "date",
  values_to = "cases") %>%
select(-c(Lat, Long))

# Repeat for deaths

global_deaths <- global_deaths %>%
  pivot_longer(cols = 
-c('Province/State', 'Country/Region', Lat, Long),
  names_to = "date",
  values_to = "deaths") %>%
select(-c(Lat, Long))

global_deaths <- global_deaths %>%
  rename(Province_State = 'Province/State',
         Country_Region = 'Country/Region')

global_cases <- global_cases %>%
  rename(Province_State = 'Province/State',
         Country_Region = 'Country/Region')

# Combine global_cases and global_deaths

global <- global_cases %>%
  full_join(global_deaths) %>%
  mutate(date = mdy(date))
```

Let's view the global dataset and look for any issues.

```{r summary}
summary(global)
```

Many rows have 0 cases. We will filter these out to only look at dates with reported cases.

```{r cleaning}
# Filtering
global <- global %>% filter(cases > 0)

```


Now we will look at US_cases and US_deaths and tidy the dataset. We will join these two datasets together to make US dataset.

```{r US_tidy, message = FALSE}
# Checking pivot
US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases")

US_cases <- US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))


# Same process with US_deaths

US_deaths <- US_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

# Joining US data
US <- US_cases %>%
  full_join(US_deaths)
```

To compare the US data to the global data, we want to have a population column and a combined key column. We will add those now.

```{r global_new, message = FALSE}
# Creating Combined Key with Province_State and Country_Region
global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)

# Adding population data
lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

global <- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)

```

At this point, we are done with tidying and organizing our data. We will move on to visualizing the data and basic analysis.

## Visual Analysis

```{r US_visual_analysis, message = FALSE}
# Create new dataset of US by state
US_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

# create US totals 
US_totals <- US_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

```


Create basic visual of US data comparing cases reported to the deaths reported.

```{r US_visual_US}

US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) + 
  geom_line(aes(color = "cases")) + 
  geom_point(aes(color = "cases")) + 
  geom_line(aes(y = deaths, color = "deaths")) + 
  geom_point(aes(y = deaths, color = "deaths")) + 
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US", y = NULL)


```


Lets narrow our search down to the state I reside in, Louisiana.

```{r US_visual_LA, message = FALSE}

state <- "Louisiana"
US_by_state %>%
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) + 
  geom_line(aes(color = "cases")) + 
  geom_point(aes(color = "cases")) + 
  geom_line(aes(y = deaths, color = "deaths")) + 
  geom_point(aes(y = deaths, color = "deaths")) + 
  scale_y_log10() + 
  theme(legend.position="bottom", axis.text.x = element_text(angle = 90)) + 
  labs(title = str_c("COVID19 in ", state), y = NULL)

# Look at recent date
max(US_totals$date)

# maximum deaths
max(US_totals$deaths)


```

## New Cases and New Deaths

After looking at the graph, I have some new questions. Has the number of cases and deaths leveled off?

To answer this, we transform our data and add new variables.

```{r level_analysis, message = FALSE}
# Add new_cases and new_deaths columns to dataset
US_by_state <- US_by_state %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
US_totals <- US_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

tail(US_totals %>% select(new_cases, new_deaths, everything()))

# Now lets graph the new cases and new deaths

US_totals %>%
  #filter(cases > 0) %>%
  ggplot(aes(x = date, y = new_cases)) + 
  geom_line(aes(color = "new_cases")) + 
  geom_point(aes(color = "new_cases")) + 
  geom_line(aes(y = new_deaths, color = "new_deaths")) + 
  geom_point(aes(y = new_deaths, color = "new_deaths")) + 
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "New COVID19 in US", y = NULL)


```

From the graph, we see that new covid 19 cases and deaths resulting from that have flatlined and stabilized over time. Lets look at the state of Louisiana to see if this trend is similar to the US in total.

```{r level_analysis_LA, message = FALSE}
state <- "Louisiana"

US_by_state %>%
  filter(Province_State == state) %>%
  #filter(cases > 0) %>%
  ggplot(aes(x = date, y = new_cases)) + 
  geom_line(aes(color = "new_cases")) + 
  geom_point(aes(color = "new_cases")) + 
  geom_line(aes(y = new_deaths, color = "new_deaths")) + 
  geom_point(aes(y = new_deaths, color = "new_deaths")) + 
  scale_y_log10() + 
  theme(legend.position="bottom", axis.text.x = element_text(angle = 90)) + 
  labs(title = str_c("New COVID19 in ", state), y = NULL)

```

The Louisiana dataset is incredibly noisy. Does this have something to do with a lack of datapoints? Lets look at a state with higher population. I have chosen to look at California, the state with the highest population.

```{r level_analysis_CA, message = FALSE}

state <- "California"

US_by_state %>%
  filter(Province_State == state) %>%
  ggplot(aes(x = date, y = new_cases)) + 
  geom_line(aes(color = "new_cases")) + 
  geom_point(aes(color = "new_cases")) + 
  geom_line(aes(y = new_deaths, color = "new_deaths")) + 
  geom_point(aes(y = new_deaths, color = "new_deaths")) + 
  scale_y_log10() + 
  theme(legend.position="bottom", axis.text.x = element_text(angle = 90)) + 
  labs(title = str_c("New COVID19 in ", state), y = NULL)

```

From the California graph, we see a similar trend with less noise than Louisiana. I believe that it is true that the noise from the Louisiana graph was caused by a lack of data.

## Best and Worst States

We want to know the worst and best states. How do we measure this? Lets do more analysis!

```{r best_worst_analysis}

# Best state
US_state_totals <- US_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases), population = max(Population),
            cases_per_thou = 1000 * cases / population,
            deaths_per_thou = 1000 * deaths / population) %>%
  filter(cases > 0, population > 0)

# Slice on US_state_totals
#US_state_totals %>%
  #slice_min(deaths_per_thou, n = 10)

US_state_totals %>%
  slice_min(deaths_per_thou, n = 10) %>%
  select(deaths_per_thou, cases_per_thou, everything())

# Worst state
US_state_totals %>%
  slice_max(deaths_per_thou, n = 10) %>%
  select(deaths_per_thou, cases_per_thou, everything())

```

## Modeling Data

Originally, we used a linear model to see if a variable is statistically significant. Alternatively, I have chosen to use a poisson regression model to predict deaths_per_thou.

```{r prepare_poisson, message = FALSE}

library(MASS)

# Prepare data
US_state_totals <- US_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), 
            cases = max(cases), 
            population = max(Population),
            cases_per_thou = 1000 * cases / population,
            deaths_per_thou = 1000 * deaths / population) %>%
  filter(cases > 0, population > 0)

```

Once our data is prepared, we will create the poisson regression model.

```{r poisson}

poisson_mod <- glm(deaths ~ cases_per_thou + population,
                   data = US_state_totals,
                   family = poisson(link = "log"))

summary(poisson_mod)

```

From our summary, we can see that both cases_per_thou and population have statistical significant influence on deaths_per_thou. In particular, cases_per_thou is the most impactful variable.



Using our model, we will add predictions to the dataset and visualize our actual deaths vs predicted deaths

```{r prediction}

# add predictions

US_tot_pred <- US_state_totals %>%
  mutate(pred_deaths = predict(poisson_mod, type = "response"))

# Visual of actual vs predicted

ggplot(US_tot_pred) + 
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue", alpha = 0.7) + 
  geom_point(aes(x = cases_per_thou, y = pred_deaths), color = "red", alpha = 0.7) +
  labs(title = "Actual vs Predicted Deaths per Thousand",
       x = "Cases per Thousand",
       y = "Deaths per Thousand") +
  theme_minimal()
  

```


Looking at the graph, while the prediction follows the general trend, I wouldn't say the predictive model is accurate. Other factors seem to influence deaths_per_thou.



## Bias

There is always bias involved with choosing a particular topic. In particular, my decision on what variables to focus on invovled bias. There is also bias in choosing to use poisson regression as opposed to other model types.

## Conclusion

Our analysis of COVID 19 data revealed several insights into the pandemic's impact across the US. Initially, we found that the new cases and new deaths stabilized over time. Using our poisson regression model, we gained a better understanding of the factors that influenced death rates across the US. The model also highlighted the complexity of pandemic impact, showing that relying on cases and population alone is not sufficient to predict the number of deaths.

